---
title: "Input - Output analysis"
format: html
execute:
  echo: true
  messages: false
  warning: false
editor: source
---

### Matrices

In our particular case, we will represent subsets of the input output table as matrices, and apply linear algebra operations to them.

This forces us to have to work in more than one dimension, in R we can do that with **matrices** and **arrays**.

A matrix is nothing more than a two-dimensional structure that allows us to store homogeneous data. To construct a matrix we use the `matrix()` function

```{r matrices}

vector_6 <- seq(8)

vector_6

A <- matrix(vector_6)

A


```

We have a one-column vector. What if we want more than one dimension?

```{r matrix2}

A <- matrix(vector_6, nrow = 2)

A


```

```{r matrices3}

A <- matrix(vector_6, ncol = 2)

A

```


```{r matrices4}

A <- matrix(vector_6, nrow = 4, ncol = 2)

A

```


We notice that the matrix is *filled* by columns, we can change that behaviour

```{r matrices5}

A <- matrix(vector_6, nrow = 4, ncol = 2, byrow = TRUE)

A

```

Just as we had the name attribute for vectors, we can also name the rows and columns of a matrix. To inspect the row names of a matrix we use `rownames()`, and the row names `colnames()`.


```{r arrays6}

rownames(A)
colnames(A)

```


NULL indicates when an expression or function results in an undefined value. In this case it indicates that the array has no names. Let's assign names to it by associating a vector of characters

```{r arrays7}

rownames(A) <- c("row_1", "row_2", "row_3", "row_4")
colnames(A) <- c("column_1", "column_2")

rownames(A)
colnames(A)

A

```

``dimnames()` is another way to inspect/assign the names of an array. 

```{r arrays8}

dimnames(A)

```

Is there anything strange about this structure? There are *lists*, a topic that we will not be covered in this course.

Accessing the elements of an array is very similar to the case of vectors, we just need to differentiate between dimensions with a comma `,`.

```{r arrays9}

A

A[1,1]

A[3,2]

A["row_2", "column_2"]

A[1:2, 1]

A[c("row_1", "row_4"), 2]

```

One operation that is widely used in practice is to create a new *table* from another by binding rows and columns. This is done with the `rbind()` and `cbind()` functions.

If we want to add columns based on another matrix:


```{r matriz41}

set.seed(125)

matrix_A <- matrix(data = runif(n = 8, min = 10, max = 50), nrow = 2, ncol = 4)

matrix_B <- matrix(data = runif(n = 2, min = 10, max = 50), nrow = 2, ncol = 1)

matrix_A

matrix_B

```

We combine both matrices

```{r matriz42}

matrix_AB <- cbind(matrix_A, matrix_B)

matrix_AB

```


If we want to add rows based on another matrix:

```{r matriz51}

matrix_C <- matrix(data = runif(n = 5, min = 10, max = 50), nrow = 1, ncol = 5)

matrix_C

matrix_ABC <- rbind(matrix_AB, matrix_C)

matrix_ABC

```

As mentioned before, matrices are objects that can be manipulated as matrices in the sense of linear algebra. For example, we can calculate the transpose or the inverse of a matrix

Transpose

```{r matrix6}

A

t(A)

```

Inverse

```{r matrix6_5}

A

solve(A[1:2,1:2])

```

Diagonal

```{r matrix7}

vector_diagonal <- c(1,2,3,4)

diag(vector_diagonal)

```

In this case

```{r matrix8}

diag(4)

```

The `diag(n)` command generates a square diagonal matrix of ones whit dimension n

The `%*%` operator performs products of matrices, as we will see in the practical examples

::: {.callout-note}

Use the help to determine what the following linear algebra functions do

`det()`, `svd()`, `eigen()`, `qr()`, `chol()`

:::


### Reading data in R

In this presentation we will focus on two formats: Excel and csv.

#### csv files

The R base function for reading csv files is `read.csv()`.

We need to provide the function with the file's location path.

```{r csv01}
#| eval: false

mtcars_csv <- read.csv(file = "mtcars.csv")

head(mtcars_csv)

```

Inspecting the function, we can find additional arguments. Let us explore them using the tab button.

In practice, we typically use the arguments  `header`, `sep` and `dec`

#### xlsx files

Base R does not have a function to read Excel files, so we must rely on a library. The most common choice is the `readxl` package. 

Once we load the library, we will use the function `read_excel()`

```{r xlsx01}
#| eval: false

#install.packages("writexl")
library(readxl)

mtcars_excel <- read_excel("mtcars.xlsx")

head(mtcars_excel)

```

In the practice section, we will use some of the other arguments of the read_excel() function. It's worth mentioning the skip argument, which is used if we don't wish to read the first n lines of an Excel file.

### Writing data in R

Unsurprisingly, we write data from R using the functions `write.csv()` and `write_xlsx()`

We must provide both functions with the object (data frame) we wish to save and the file path.

```{r xlsx02}
#| eval: false

write.csv(x = airquality, file = "my_airquality.csv")

write_xlsx(x = airquality, path = "my_airquality.xlsx")

```


# Input Output Analysis

We will use as an example a input output table of Belgium with 10 sectors

## Read the data

We have our input-output table in Excel format. In order to manipulate this data in R, we must have an R object in memory.

When we read data from an Excel or csv file, this object is loaded as a data frame.

```{r io01}

#install.packages(readxlsb)
library(readxlsb)

BE_Z <- read_xlsb(path = "./data/FIGARO_ICIOI_23ed_2021.xlsb",
                  sheet = "BE",
                  range = "C12:L22")

BE_Ytemp <- read_xlsb(path = "./data/FIGARO_ICIOI_23ed_2021.xlsb",
                      sheet = "BE",
                      range = "M12:R22")

BE_X <- read_xlsb(path = "./data/FIGARO_ICIOI_23ed_2021.xlsb",
                  sheet = "BE",
                  range = "S12:S22")

```


## Technical coefficients matrix

Let's revisit the representation of interindustry production relations using the interindustry sales matrix $Z$ and the final demand vector $y$ as a system of linear equations in matrix form:

$$x=Zi+y$$
In this equation, $Z$ represents the interindustry sales matrix, $i$ is a column vector of ones, and $y$ is a vector representing the final demand by industry.

We can define the technical coefficient $a_{ij}$ as:

$$a_{ij}=\frac{z_{ij}}{x_j}$$

Here, $z_{ij}$ represents the value of interindustry sales between industries $i$ and $j$, and $x_j$ represents the value of production in industry $j$.

Using the technical coefficient definition, we can rewrite $z_{ij}$ as $z_{ij}=a_{ij}x_j$, which can be represented in matrix notation as:

$$x=Ax+y$$

Where $A$ represents the technical coefficient matrix.

Once we have our input-output data loaded in R, the next step is to construct our $Z$ and $A$ matrices. While our data may initially be loaded as data frames in R, it's important to note that we need to convert them to matrices in order to perform matrix algebra operations. Therefore, our first task is to convert the data frames to matrices.


```{r io2}
#| output: false

Z <- as.matrix(BE_Z)

```

We add names to the matrix rows

```{r io21}
#| output: false

rownames(Z) <- colnames(Z)

```

Next, we repeat this process with the output vector. Specifically, we need to extract the output column from the `BE_X` data frame and transform it into a vector, as opposed to a one-dimensional matrix. It's important to note the distinction, as using a vector in this context is more appropriate for the calculations we will be performing.


```{r io22}
#| output: false

x <- as.vector(BE_X$X)

```

We now have all the necessary elements to compute the technical coefficients matrix.

```{r io23}
#| output: false

A = Z %*% diag(1 / x) 

colnames(A) <- rownames(A)

```


## Leontief inverse

From the equation $x=Ax+y$, we can derive an expression that informs us of the output required to meet a given exogenous level of final demand:


$$x=(I-A)^{-1}y$$

where $(I-A)^{-1}$ is known as the *Leontief inverse* or the total requirements matrix.

We compute the Leontief inverse in R:


```{r io3}
#| output: false

#Number of sectors
n <- length(x) 

#Identy matrix
I <- diag(n) 

#Leontief inverse
B <- solve(I - A) 

```

## Multipliers

Applying the change operator $\Delta$ to both sides of the expression $x=(I-A)^{-1}y$ we obtain

$$\Delta x=(I-A)^{-1}\Delta y$$
Next, let's consider the relationship between a unit change in the demand vector $\Delta y$ and the resulting change in production $\Delta x$, which can be expressed as $(I-A)^{-1}$.
Let's define $B=(I-A)^{-1}$.

The matrix $B$ enables us to calculate the increase in the production of each sector in response to changes in exogenous factors, such as variations in final demand


```{r io4}
#| output: false

#Number of sectors
n <- length(x) 

#Identy matrix
I <- diag(n) 

#Leontief matrix
B <- solve(I - A) 

```

### Interpretation

For instance, let's compare the interpretation of the elements of $A$ and $B$, such as the elements in the first and second rows and the first column of matrix $A$

```{r io5}

A[1, 1]
A[2, 1]

```

The element $A_{1,1}$ indicates that in order to produce 1 unit of output in the agriculture, forestry, and fishing industry, `r A[1, 1]` units of the agriculture, forestry, and fishing industry are required.

Similarly, the element $A_{2,1}$ tells us that to produce 1 unit of output in the agriculture, forestry, and fishing industry, `r A[2, 1]` units of the mining and quarrying industry are needed.

By summing the columns of matrix $A$, we can obtain the *direct effect* of a unitary change in the demand for the product in a specific industry

```{r io52}

direct <- colSums(A, na.rm = TRUE)

direct

direct[which.max(direct)]
sort(direct, decreasing = TRUE)

```

These numbers do not account for the feedback loops generated by the interdependence between sectors.

The matrix $B$ incorporates the *indirect* effect stemming from the fact that industries supplying the agriculture sector, for instance, also require inputs from other industries.

By summing the columns of matrix $B$, we obtain the *direct + indirect effects* of a unitary change in the demand for the product in a specific industry.

For example, let's consider the elements in the first and second rows and the first column of matrix $B$


```{r io53}

B[1, 1]
B[2, 1]

```

The element $B_{1,1}$ indicates that in order to meet a 1 unit increase in demand for agriculture, forestry, and fishing industry products, `r B[1, 1]` units of agriculture, forestry, and fishing industry input are required. This includes the additional unit itself, as well as `r B[1, 1] - 1` units needed as input for other sectors.

Similarly, the element $B_{2,1}$ shows that to satisfy a 1 unit increase in demand for agriculture, forestry, and fishing industry products, `r B[2, 1]` units of mining and quarrying industry input are necessary.

By calculating the column sums of matrix $B$, we can determine the *multiplier* for industry $j$, which represents the total value of production in all industries required for all stages of production in order to produce one unit of product $j$ for final use in the economy

```{r io54}

total <- colSums(B, na.rm = TRUE)

total[which.max(total)]
sort(total, decreasing = TRUE)

```

